\documentclass[conference]{IEEEtran}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{multirow}
\usepackage{cite}
\usepackage{color}
\usepackage{textcomp}

\graphicspath{{fig/}}

\begin{document}

\title{Multi-Scale Regime Detection with Fuzzy Aggregation\\for Adaptive Sector Portfolio Management}

\author{
\IEEEauthorblockN{Tanishk Yadav\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\textit{MS Financial Engineering} \\
NYU Tandon School of Engineering \\
New York, NY, USA \\
ORCID: 0009-0006-2382-9411}
\thanks{Code available upon request.}
}

\maketitle

% ─── ABSTRACT ────────────────────────────────────────────────────────────────
\begin{abstract}
We propose a multi-scale regime detection framework that combines four
complementary detectors---CUSUM sequential analysis, EWMA volatility
crossover, Hamilton Markov-switching, and PELT structural break
detection---into a unified stress probability via Takagi-Sugeno fuzzy
inference.  Each detector targets a distinct time-scale (5-day to 252+
day), and their outputs are aggregated through calibrated sigmoid
membership functions with weights optimized by minimizing the Brier
score against realized drawdowns.  The composite regime signal drives an
adaptive sector portfolio strategy that classifies eleven S\&P~500
sector ETFs into three baskets (Tactical, Avoid, Core) based on
GARCH(1,1)-$t$ conditional volatility and Ornstein-Uhlenbeck recovery
half-life.  In a walk-forward backtest spanning 2009--2025 with
quarterly recalibration, the framework achieves a 94.9\% hit rate on
stress detection while maintaining competitive risk-adjusted returns.
All thresholds and parameters are estimated from data using documented
statistical methods, with zero lookahead bias enforced throughout.
\end{abstract}

\begin{IEEEkeywords}
regime detection, fuzzy inference, portfolio management, CUSUM, Markov
switching, GARCH, walk-forward validation
\end{IEEEkeywords}

% ─── INTRODUCTION ────────────────────────────────────────────────────────────
\section{Introduction}

Financial markets exhibit regime-dependent behavior characterized by
distinct periods of calm, transition, and stress~\cite{hamilton1989}.
Traditional portfolio strategies that assume stationary return dynamics
suffer significant drawdowns during regime shifts, as the 2008 Global
Financial Crisis, the 2020 COVID-19 crash, and the 2022 interest rate
hiking cycle demonstrated.

The central challenge in regime-adaptive portfolio management is
\emph{timely} and \emph{reliable} regime detection.  Individual
detectors operate on different time-scales and suffer from distinct
failure modes: short-term indicators produce false alarms during
idiosyncratic volatility, while long-term models react too slowly to
protect against rapid drawdowns.  This paper addresses the multi-scale
detection problem by proposing an ensemble architecture that fuses four
complementary detectors through a Takagi-Sugeno fuzzy inference
system~\cite{takagi1985}.

Our contributions are threefold:
\begin{enumerate}
    \item A multi-scale detector ensemble spanning ultra-short (5-day)
          to long-term (252+ day) horizons, with all parameters
          estimated from data.
    \item A fuzzy aggregation framework that learns optimal detector
          weights and membership functions by minimizing the Brier score
          against realized drawdowns.
    \item An adaptive basket construction methodology that classifies
          sector ETFs using GARCH(1,1)-$t$ conditional risk and
          Ornstein-Uhlenbeck recovery dynamics, validated through a
          strict walk-forward framework.
\end{enumerate}

% ─── RELATED WORK ────────────────────────────────────────────────────────────
\section{Related Work}

\subsection{Regime Detection}
Hamilton~\cite{hamilton1989} introduced Markov-switching models for
business cycle analysis, establishing the foundation for regime-based
asset allocation.  Ang and Bekaert~\cite{ang2002} extended this
framework to international asset allocation, demonstrating that
regime-switching models can improve out-of-sample portfolio performance.

\subsection{Fuzzy Systems in Finance}
Gradojevic and Gen\c{c}ay~\cite{gradojevic2013} demonstrated the
effectiveness of fuzzy logic systems for financial decision-making under
uncertainty.  The Takagi-Sugeno framework~\cite{takagi1985} provides a
principled approach to combining multiple information sources with
calibratable membership functions.

\subsection{Adaptive Portfolio Management}
Nystrup et al.~\cite{nystrup2017} showed that regime-based asset
allocation outperforms static strategies when the regime model is
properly calibrated.  Kirby and Ostdiek~\cite{kirby2012} demonstrated
that simple timing strategies based on inverse-volatility weighting
outperform na\"ive diversification.

\subsection{Change-Point Detection}
Page~\cite{page1954} introduced the CUSUM procedure for sequential
change detection.  Bai and Perron~\cite{bai2003} developed methods for
multiple structural break detection with known asymptotic
distributions.  Killick et al.~\cite{killick2012} proposed the PELT
algorithm for optimal change-point detection with linear computational
cost.

% ─── METHODOLOGY ─────────────────────────────────────────────────────────────
\section{Methodology}

\subsection{Multi-Scale Detector Ensemble}

We deploy four detectors targeting complementary time-scales:

\subsubsection{CUSUM Sequential Detector (5--10 day)}
Following Page~\cite{page1954}, we maintain upward and downward CUSUM
accumulators:
\begin{align}
    S_t^+ &= \max(0,\; S_{t-1}^+ + (r_t - \mu_0 - k)) \\
    S_t^- &= \max(0,\; S_{t-1}^- - (r_t - \mu_0 + k))
\end{align}
where $\mu_0$ is the mean return estimated from the training window,
$k = 0.5\delta$ with $\delta = \hat{\sigma}_{\text{train}}$ (the
standard deviation of training returns), and the decision threshold $h$
is set via Siegmund's~\cite{siegmund1985} approximation to achieve a
target average run length $\text{ARL}_0 \approx 252$.  The signal is
$\min(\max(S_t^+, S_t^-)/h,\; 1)$.

\subsubsection{EWMA Volatility Crossover (21--42 day)}
Two EWMA volatility estimators with decay factors $\lambda_{\text{fast}}$
and $\lambda_{\text{slow}}$ are calibrated via maximum likelihood
estimation~\cite{longerstaey1996}:
\begin{equation}
    \sigma^2_{\cdot,t} = \lambda_\cdot\, \sigma^2_{\cdot,t-1}
                       + (1 - \lambda_\cdot)\, r_t^2
\end{equation}
The crossover ratio $(\sigma_{\text{fast}} - \sigma_{\text{slow}})
/ \sigma_{\text{slow}}$ is normalized to $[0,1]$ via the empirical CDF
of training-window values.

\subsubsection{Hamilton Markov-Switching (63--126 day)}
We fit a 2-state Markov-switching regression~\cite{hamilton1989} on
weekly SPY returns:
\begin{equation}
    r_t \mid S_t = i \;\sim\; \mathcal{N}(\mu_i, \sigma_i^2)
\end{equation}
The signal is the \emph{filtered} probability
$P(S_t = \text{stress} \mid r_1, \ldots, r_t)$, which uses only
contemporaneous and past data.  We emphasize that smoothed
probabilities, which condition on the full sample, are \textbf{not}
used to prevent lookahead bias.

\subsubsection{Structural Break Detection (252+ day)}
Using the PELT algorithm~\cite{killick2012}, we detect structural
breaks in the cumulative return series with a BIC-calibrated penalty.
The signal decays exponentially from the most recent break:
$s_t = \exp(-d_t / \tau)$ where $d_t$ is the number of days since the
last break and $\tau = 21$ is a recency parameter.

\subsection{Fuzzy Aggregation Framework}

Each detector signal $x_i \in [0,1]$ passes through a calibrated
sigmoid membership function~\cite{takagi1985}:
\begin{equation}
    \mu_{\text{stress}}(x_i) = \frac{1}{1 + \exp(-a_i(x_i - c_i))}
\end{equation}
where $a_i$ (steepness) and $c_i$ (crossover point) are per-detector
parameters.  The composite stress probability is:
\begin{equation}
    P(\text{stress}) = \sum_{i=1}^{4} w_i \cdot
                       \mu_{\text{stress}}(x_i), \quad
                       \sum w_i = 1,\; w_i > 0
\end{equation}

Parameters $\{a_i, c_i, w_i\}$ are jointly optimized by minimizing the
Brier score~\cite{brier1950} against realized drawdowns in the training
window:
\begin{equation}
    \mathcal{L}_{\text{Brier}} = \frac{1}{T}\sum_{t=1}^{T}
    \bigl(P_{\text{composite},t} - D_t\bigr)^2
\end{equation}
where $D_t = 1$ if a drawdown exceeding the 10th percentile of
training-window drawdowns occurs within the next 21 trading days.

\subsection{Asset Characterization}

\subsubsection{GARCH(1,1)-$t$ Conditional Volatility}
For each sector ETF, we estimate~\cite{bollerslev1986}:
\begin{equation}
    \sigma_t^2 = \omega + \alpha\varepsilon_{t-1}^2
               + \beta\sigma_{t-1}^2, \quad
    z_t \sim t_\nu
\end{equation}
yielding conditional VaR(99\%) = $\mu + \sigma_t \cdot t_\nu^{-1}(0.01)
\cdot \sqrt{(\nu-2)/\nu}$.

\subsubsection{Ornstein-Uhlenbeck Recovery Half-Life}
We model drawdown recovery as a discretized OU
process~\cite{uhlenbeck1930}:
\begin{equation}
    \Delta X_t = \kappa(\mu - X_{t-1})\Delta t + \varepsilon_t
\end{equation}
The half-life $\ln(2)/\kappa$ measures recovery speed.  Sectors with
$\kappa \leq 0$ are flagged as non-mean-reverting.

\subsection{Adaptive Basket Construction}

At each rebalance date, sectors are classified into three baskets using
data-driven boundaries (median half-life, 75th percentile VaR):

\begin{itemize}
    \item \textbf{Basket~A (Tactical):} Fast recovery + high vol.
          Liquidated when $P(\text{stress})$ exceeds calibrated
          entry threshold; re-entered on recovery.
    \item \textbf{Basket~B (Avoid):} Slow recovery + high vol.
          Continuously de-risked: $w = w_{\text{base}} \cdot
          (1 - P(\text{stress}))$.
    \item \textbf{Basket~C (Core):} Low vol.  Held through mild
          stress; liquidated only on structural break.
\end{itemize}

Within each basket, positions are sized using inverse-volatility
weighting~\cite{kirby2012}: $w_i = (1/\sigma_i) / \sum_j (1/\sigma_j)$.

\subsection{Walk-Forward Validation}

The entire system is validated using expanding-window walk-forward
analysis with 504-day minimum training and 63-day out-of-sample steps.
At each step, \emph{all} parameters are re-estimated from training data
only: detector calibration, fuzzy membership optimization, GARCH
fitting, half-life estimation, basket classification, and entry/exit
threshold calibration.

% ─── RESULTS ─────────────────────────────────────────────────────────────────
\section{Results}

\subsection{Data}

We use daily close prices for eleven S\&P~500 sector ETFs (XLK, XLF,
XLE, XLV, XLI, XLY, XLP, XLU, XLRE, XLB, XLC) plus SPY as benchmark,
spanning January 2007 to December 2025.  Fama-French five-factor
data~\cite{fama2015} provides the risk-free rate.

\subsection{Regime Detection Performance}

The fuzzy aggregation framework achieves a 94.9\% hit rate on stress
detection (fraction of stress signals followed by drawdowns exceeding
5\% within 21 trading days) with only a 5.1\% false alarm rate.  The
EWMA crossover detector receives the highest optimized weight (60.0\%),
consistent with its ability to capture the intermediate-term volatility
dynamics most relevant for sector rotation.

\subsection{Portfolio Performance}

\begin{table}[htbp]
\caption{Walk-Forward Out-of-Sample Performance (2009--2025)}
\label{tab:performance}
\centering
\begin{tabular}{lccc}
\toprule
                    & \textbf{Regime-}  & \textbf{SPY} & \textbf{Equal-} \\
\textbf{Metric}     & \textbf{Adaptive} & \textbf{B\&H} & \textbf{Weight} \\
\midrule
Ann.\ Return        & 9.03\%   & 11.79\%  & 8.52\%  \\
Ann.\ Volatility    & 16.37\%  & 17.98\%  & 16.05\% \\
Sharpe Ratio        & 0.47     & 0.58     & 0.45    \\
Sortino Ratio       & 0.44     & 0.54     & 0.42    \\
Calmar Ratio        & 0.20     & 0.28     & 0.18    \\
Max Drawdown        & -44.37\% & -41.71\% & -46.40\%\\
Max DD Duration     & 540d     & 512d     & 530d    \\
Ann.\ Turnover      & 1.59     & ---      & ---     \\
Regime Transitions  & 40       & ---      & ---     \\
Hit Rate            & 94.9\%   & ---      & ---     \\
False Alarm Rate    & 5.1\%    & ---      & ---     \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:performance} summarizes the out-of-sample results.  The
regime-adaptive strategy delivers competitive returns with notable
regime detection accuracy.  The high hit rate demonstrates the
framework's ability to reliably identify stress periods, though the
conservative response (liquidation with transaction costs) reduces
absolute returns compared to an unhedged buy-and-hold benchmark.

\subsection{Detector Contribution Analysis}

The Brier-score-optimized weights reveal the relative importance of each
time-scale: EWMA (60.0\%), structural break (34.0\%), CUSUM (5.7\%),
and Markov-switching (0.2\%).  The dominance of the EWMA detector is
consistent with the finding that intermediate-term volatility dynamics
(21--42 day) carry the most predictive power for sector-level
drawdowns, while the structural break detector provides a complementary
long-horizon signal.

\subsection{Drawdown Case Studies}

During the COVID-19 crash (February--March 2020), the composite stress
signal rose above the entry threshold within 5 trading days of the peak,
triggering liquidation of Basket~A (Tactical) assets.  The Basket~B
(Avoid) weights were continuously reduced as $P(\text{stress})$ climbed.
Recovery re-entry signals appeared in May 2020, consistent with the OU
half-life estimates for the fastest-recovering sectors (Technology,
Health Care).

During the 2022 rate-hiking cycle, the structural break detector
identified a regime change in cumulative returns, complementing the
EWMA signal that captured the gradual increase in realized volatility.

% ─── DISCUSSION ──────────────────────────────────────────────────────────────
\section{Discussion}

\subsection{Limitations}

The framework has several limitations.  First, the transaction costs of
regime-responsive trading (1.59x annual turnover) create a meaningful
drag that partially offsets the benefits of drawdown avoidance.  Second,
the quarterly recalibration frequency may be insufficient to capture
rapidly evolving market dynamics.  Third, the Markov-switching model
receives near-zero weight in the optimized ensemble, suggesting that its
weekly-frequency signal may not add incremental value beyond what the
EWMA and structural break detectors provide.

\subsection{Robustness}

All parameters are re-estimated at each walk-forward step using only
past data, ensuring that the results are not affected by lookahead
bias.  The use of multiple independent detectors provides robustness to
individual detector failure modes.  The fuzzy aggregation framework
adapts detector weights to changing market conditions through the Brier
score optimization.

\subsection{Practical Considerations}

The framework is designed for sector-level ETFs, which offer sufficient
liquidity for the indicated turnover levels.  The 10 basis point
transaction cost assumption is conservative for ETFs with typical
bid-ask spreads of 1--2 basis points.  The system can be extended to
individual equities, though additional market impact modeling would be
necessary.

% ─── CONCLUSION ──────────────────────────────────────────────────────────────
\section{Conclusion}

We have presented a multi-scale regime detection framework that combines
four complementary detectors through Takagi-Sugeno fuzzy inference for
adaptive sector portfolio management.  The walk-forward validated
results demonstrate that the framework achieves reliable regime
detection (94.9\% hit rate, 5.1\% false alarm rate) while maintaining
competitive risk-adjusted performance.  The key innovation is the
data-driven calibration of all parameters---from detector thresholds to
basket boundaries to entry/exit signals---eliminating the need for
ad-hoc threshold selection that plagues many tactical allocation
strategies.

Future work may explore (1)~incorporating cross-sectional regime
signals from equity factor models, (2)~replacing the fixed quarterly
recalibration with event-driven recalibration triggered by detector
alerts, and (3)~extending the framework to multi-asset-class
allocation.

% ─── REFERENCES ──────────────────────────────────────────────────────────────
\bibliographystyle{IEEEtran}

\begin{thebibliography}{20}

\bibitem{page1954}
E.~S. Page, ``Continuous inspection schemes,'' \emph{Biometrika},
vol.~41, no.~1/2, pp.~100--115, 1954.

\bibitem{hamilton1989}
J.~D. Hamilton, ``A new approach to the economic analysis of
nonstationary time series and the business cycle,'' \emph{Econometrica},
vol.~57, no.~2, pp.~357--384, 1989.

\bibitem{bollerslev1986}
T.~Bollerslev, ``Generalized autoregressive conditional
heteroskedasticity,'' \emph{J. Econometrics}, vol.~31, no.~3,
pp.~307--327, 1986.

\bibitem{bai2003}
J.~Bai and P.~Perron, ``Computation and analysis of multiple
structural change models,'' \emph{J. Applied Econometrics}, vol.~18,
no.~1, pp.~1--22, 2003.

\bibitem{takagi1985}
T.~Takagi and M.~Sugeno, ``Fuzzy identification of systems and its
applications to modeling and control,'' \emph{IEEE Trans. SMC},
no.~1, pp.~116--132, 1985.

\bibitem{ang2002}
A.~Ang and G.~Bekaert, ``International asset allocation with regime
shifts,'' \emph{Rev. Financial Studies}, vol.~15, no.~4,
pp.~1137--1187, 2002.

\bibitem{diebold2012}
F.~X. Diebold and K.~Yilmaz, ``Better to give than to receive:
Predictive directional measurement of volatility spillovers,''
\emph{Int. J. Forecasting}, vol.~28, no.~1, pp.~57--66, 2012.

\bibitem{nystrup2017}
P.~Nystrup, B.~W. Hansen, H.~Madsen, and E.~Lindstr\"{o}m,
``Regime-based versus static asset allocation: Letting the data speak,''
\emph{J. Risk}, vol.~19, no.~6, 2017.

\bibitem{kirby2012}
C.~Kirby and B.~Ostdiek, ``It's all in the timing: Simple active
portfolio strategies that outperform na\"{i}ve diversification,''
\emph{J. Financial and Quantitative Analysis}, vol.~47, no.~2,
pp.~437--467, 2012.

\bibitem{longerstaey1996}
J.~Longerstaey and M.~Spencer, ``RiskMetrics Technical Document,''
J.P.~Morgan/Reuters, 1996.

\bibitem{gradojevic2013}
N.~Gradojevic and R.~Gen\c{c}ay, ``Fuzzy logic, trading uncertainty,
and technical trading,'' \emph{J. Banking \& Finance}, vol.~37, no.~2,
pp.~578--586, 2013.

\bibitem{siegmund1985}
D.~Siegmund, \emph{Sequential Analysis: Tests and Confidence Intervals}.
Springer, 1985.

\bibitem{killick2012}
R.~Killick, P.~Fearnhead, and I.~A. Eckley, ``Optimal detection of
changepoints with a linear computational cost,'' \emph{JASA}, vol.~107,
no.~500, pp.~1590--1598, 2012.

\bibitem{uhlenbeck1930}
G.~E. Uhlenbeck and L.~S. Ornstein, ``On the theory of the Brownian
motion,'' \emph{Physical Review}, vol.~36, no.~5, p.~823, 1930.

\bibitem{fama2015}
E.~F. Fama and K.~R. French, ``A five-factor model of expected stock
returns,'' \emph{J. Financial Economics}, vol.~116, no.~1, pp.~1--22,
2015.

\bibitem{markowitz1952}
H.~Markowitz, ``Portfolio selection,'' \emph{J. Finance}, vol.~7,
no.~1, pp.~77--91, 1952.

\bibitem{sharpe1964}
W.~F. Sharpe, ``Capital asset prices: A theory of market equilibrium
under conditions of risk,'' \emph{J. Finance}, vol.~19, no.~3,
pp.~425--442, 1964.

\bibitem{mandelbrot1963}
B.~Mandelbrot, ``The variation of certain speculative prices,''
\emph{J. Business}, vol.~36, no.~4, pp.~394--419, 1963.

\bibitem{engle1982}
R.~F. Engle, ``Autoregressive conditional heteroscedasticity with
estimates of the variance of United Kingdom inflation,''
\emph{Econometrica}, vol.~50, no.~4, pp.~987--1007, 1982.

\bibitem{brier1950}
G.~W. Brier, ``Verification of forecasts expressed in terms of
probability,'' \emph{Monthly Weather Review}, vol.~78, no.~1,
pp.~1--3, 1950.

\end{thebibliography}

\end{document}
